---
title: "introduction to quarto"
format: html
bibliography: references.bib
---

# Introduction

## About the data

-   Water sample data from Arctic Ocean
-   Data downloaded from the Arctic Data Center: [Tweedie ADC data](https://arcticdata.io/catalog/view/doi:10.18739/A25T3FZ8X)

## Setup

```{r}
#install.packages("readr")
library(readr)
#helps read in data
library(here)
#can also check boxes in the packages tab, but then its not reproducible

```

## Read in data

```{r}
#| warning: false

bg_chem <- read_csv(here("data/BGchem2008data.csv"))
```

# Analysis

## calculate summary statistics

```{r, echo=FALSE, eval=FALSE}
#interchangeable syntax
#| echo: false
#exclude source code in the output
#| eval: false
#exclude analysis in the output

colnames(bg_chem)
#print column names

str(bg_chem)
#show the data structure

head(bg_chem)

summary(bg_chem)
#summary of each column

unique(bg_chem$Target_Depth)

```

## calculate mean Redfield ratio

```{r}
# calculate some means
nitrate<-mean(bg_chem$NO3)
#assign value
nitrite<-mean(bg_chem$NO2)
amm<-mean(bg_chem$NH4)
phos<-mean(bg_chem$P)
```

## plot Redfield ratio

```{r}
ratio <- (nitrate + nitrite + amm) / phos
```

Here is the ratio we calculated: `r round(ratio)`

#substituted the words for the value in rendering!!

# Conclusion

##reference of fonts

## hashtags is a large title

### more hastags is smaller

###### a bunch of hashtags is even smaller

*stars on both ends is italics*

**bold is two stars**

-   one front star AND space is bullet point

[@oke2020; @oke2020a]

| Col1 | Col2 | Col3 |
|------|------|------|
|      |      |      |
|      |      |      |
|      |      |      |
