---
title: "Spatial data"
format: html
editor_options: 
  chunk_output_type: console
---

```{r, message = FALSE, echo = FALSE}
library(readr)
library(sf)
library(ggplot2)
library(scales)
library(ggspatial)
library(dplyr)
```

# Load data from KNB
```{r}
knb_url <- "https://dev.nceas.ucsb.edu/knb/d1/mn/v2/object/urn%3Auuid%3Aaceaecb2-1ce0-4d41-a839-d3607d32bb58"
#read in a url as a value

# download the zip file
download.file(url = knb_url, destfile = 'shapefile_demo_data.zip')
#chose the zip file from that url

# unzip
unzip('shapefile_demo_data.zip', exdir = 'data')
#unzip the file you downloaded, exdir is where

# remove the zip file
file.remove('shapefile_demo_data.zip')
```

## Exploring the data using plot() and st_crs()
```{r}
ak_region <- read_sf('~/training_kjaer/data/ak_regions_simp.shp')

plot(ak_region)
```


```{r}
class(ak_region)

head(ak_region)

glimpse(ak_region)
```


```{r}
st_crs(ak_region)

```

### select()
```{r}
ak_regions_3338 <- ak_region %>% 
  st_transform(crs = 3338)

st_crs(ak_regions_3338)

plot(ak_regions_3338)
```

### filter()
```{r}
colnames(ak_regions_3338)

ak_regions_3338 %>% 
  select(region)
#whats sticky?
#geometry column is "sticking" to dataset but idk why

ak_regions_3338 %>%
  filter(region == "Southeast")
#13
head(ak_regions_3338)
```

## Spatial joins
How many people live in each of these Alaska region?

```{r}
library(here)
#read in population data
pop_df <- read_csv(here("data/alaska_population.csv"))

pop_4326_sf <- st_as_sf(pop_df, 
                         coords = c('lng', 'lat'), #long first, then lat
                         crs = 4326,
                        remove = F) #keep the lat long columns

head(pop_4326_sf)
```

### joining with st_join()
```{r}
pop_3338 <- st_transform(pop_4326_sf,
                         crs = 3338)

pop_joined <- st_join(pop_3338,
                      ak_regions_3338,
                      join = st_within)
#323
head(pop_joined)
```

## Calculate total population using group_by() and summarize()
```{r}
pop_region <- pop_joined %>% 
  as.data.frame() %>%  #will remove geomtry, sf -> df
  group_by(region) %>% 
  summarize(total_pop = sum(population))
#13

head(pop_region)
```

### add total pop back to shapefile
```{r}
pop_region_3338 <- left_join(ak_regions_3338,
                             pop_region,
                             by = "region"
                             )
#13

# plot of pop by region
plot(pop_region_3338["total_pop"])
```

## pop by mgmt area
```{r}
pop_mgmt_3338 <- pop_region_3338 %>% 
  group_by(mgmt_area) %>% 
  summarise(total_pop = sum(total_pop))

plot(pop_mgmt_3338["total_pop"])
```

```{r}
pop_mgmt_3338 <- pop_region_3338 %>% 
  group_by(mgmt_area) %>% 
  summarise(total_pop = sum(total_pop), do_union=F)

plot(pop_mgmt_3338["total_pop"])
```

## saving spatial object with write_sf()
```{r}
write_sf(pop_region_3338, "data/ak_regions_populations.shp")
```

# ggplot
```{r}
ggplot(pop_region_3338) +
  geom_sf(aes(fill = total_pop)) +
  labs(fill = "Total Population") + #label legend
  scale_fill_continuous(low = "khaki",
                        high = "firebrick",
                        labels = comma) #out of sci notation
```

## adding in rivers
```{r}
rivers_3338_sf <- read_sf(here("data/ak_rivers_simp.shp"))

st_crs(rivers_3338_sf)
#input is alaska albers system

ggplot() + 
  geom_sf(data = pop_region_3338,
          aes(fill = total_pop)) +
  geom_sf(data = pop_3338,
          size = 0.5) +
  geom_sf(data = rivers_3338_sf,
          aes(linewidth = StrOrder)) + # stahler stream order (larger number is larger river)
  scale_linewidth(range = c(0.05, 0.5), guide = "none") +
  labs(title = "Total Population by Alaska Region",
       fill = "Total Population") +
  scale_fill_continuous(low = "khaki",
                        high = "firebrick",
                        labels = comma) +
  theme_bw()
```

# Adding base maps with ggspatial
Pseudo-mercator (EPSG 3857)
```{r}
pop_3857_sf <- st_transform(pop_3338,
                            crs = 3857)

ggplot(data = pop_3857_sf) +
  ggspatial::annotation_map_tile(type = "osm", #free background map tiles
                                 zoom = 4, #zoom, country/state view
                                 progress = 'none') +
  geom_sf(aes(color = population),
          fill = NA) +
  scale_color_continuous(low = "darkkhaki",
                         high = "firebrick",
                         labels = comma)
```

## sf objects with leaflet
unlike ggplot, leaflet will project the data for you, given a projection system and coordinates, interactive, dragable
```{r}
# first need to save projection

epsg3338 <- leaflet::leafletCRS(
    crsClass = "L.Proj.CRS",
    code = "EPSG:3338",
    proj4def =  "+proj=aea +lat_1=55 +lat_2=65 +lat_0=50 +lon_0=-154 +x_0=0 +y_0=0 +ellps=GRS80 +towgs84=0,0,0,0,0,0,0 +units=m +no_defs",
    resolutions = 2 ^ (16:7)
)
```

## st_transform() to get unprojected coordiante system (WGS 84)
```{r}
pop_region_4326 <- pop_region_3338 %>% 
  st_transform(crs = 4326)
```

## make leaflet map
```{r}
library(leaflet)
m <- leaflet(options = leafletOptions(crs = epsg3338)) %>% 
  addPolygons(data = pop_region_4326,
              fillColor = "gray",
              weight = 1)
m
```

## make better map
```{r}
pal <- colorNumeric(palette = "Reds", domain = pop_rgn_4326$total_pop)

m <- leaflet(options = leafletOptions(crs = epsg3338)) %>%
    addPolygons(
        data = pop_rgn_4326,
        fillColor = ~ pal(total_pop),
        weight = 1,
        color = "black",
        fillOpacity = 1
    ) %>%
    addCircleMarkers(
        data = pop_4326,
        lat = ~ lat,
        lng = ~ lng,
        radius = ~ log(population / 500),
        # arbitrary scaling
        fillColor = "gray",
        fillOpacity = 1,
        weight = 0.25,
        color = "black",
        label = ~ paste0(pop_4326_sf$city, ", population ", comma(pop_4326_sf$population))
    ) %>%
    addLegend(
        position = "bottomleft",
        pal = pal,
        values = range(pop_rgn_4326$total_pop),
        title = "Total Population"
    )

m
```

