---
title: "Tidying and wrangling"
format: html
---

```{r, message=FALSE}
library(dplyr)
library(tidyr)
library(readr)
```

# Data sources

## read in our data

```{r}
#| message: FALSE
catch_og <- read_csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1")
```

## Explore

```{r}
colnames(catch_og)

head(catch_og)

summary(catch_og)

unique(catch_og$Region)

```

# Cleaning

## using the select() function

```{r}
# select() to keep the columns that we want
catch_data <- catch_og %>% 
  select(Region, Year, Chinook, Sockeye, Coho, Pink, Chum)
head(catch_data)
#6x7

catch_data1 <- catch_og %>% 
  select(-All, -notesRegCode)
head(catch_data1)
#6x7

```

## check column types
```{r}
glimpse(catch_data)
```

## Changing column contents with mutate()

```{r}

# change Chinook to numeric
catch_clean <- catch_data %>% 
  mutate(Chinook = as.numeric(Chinook))

unique(catch_data$Chinook)

is.na(catch_clean$Chinook)

# get the location of the NA value
i <- which(is.na(catch_clean$Chinook))
i

# look at the row in the data
catch_data[i,]

# use mutate to fix
catch_clean <- catch_data %>% 
  mutate(Chinook = if_else(condition = Chinook == "I", 
                            true = "1",
                            false = Chinook),
          Chinook = as.numeric(Chinook)
          )
```


## Changing file shape using pivot_longer() and pivot_wider()

```{r}
catch_long <- catch_clean %>% 
  pivot_longer(cols = -c(Region, Year),
               names_to = "species",
               values_to="catch")
head(catch_long)
#8540 of 4

catch_wide <- catch_long %>% 
  pivot_wider(names_from = species,
              values_from = catch)
#1708 of 7
  
```

## rename columns with rename()
```{r}
catch_long <- catch_long %>% 
  rename(catch_thousands = catch)
head(catch_long)

catch_long <- catch_long %>% 
  mutate(catch = catch_thousands *1000) %>% 
  select(-catch_thousands)
head(catch_long)
```

## summary statistics using group_by and summarize()
```{r}
# get average catch by region
mean_region <- catch_long %>% 
  group_by(Region) %>% 
  summarize(mean_catch = mean(catch))
head(mean_region)

# number of rows for each group
n_region <- catch_long %>%
  group_by(Region) %>% 
  summarize(n=n())
head(n_region)
```

# filtering rows using filter()
```{r}
sse_catch <- catch_long %>% 
  filter(Region == "SSE")
head(sse_catch)
#560

catch_million <- catch_long %>% 
  filter(catch >= 1000000)
head(catch_million)
#1200

chinook_sse <- catch_long %>% 
  filter(Region == "SSE" & species == "Chinook")
#112

```

## Sorting data with arrange()
```{r}
mean_region<-catch_long %>% 
  group_by(Region) %>%
  summarize(mean_catch = mean(catch)) %>% 
  arrange(mean_catch)
head(mean_region)

# desc
mean_region<-catch_long %>% 
  group_by(Region) %>%
  summarize(mean_catch = mean(catch)) %>% 
  arrange(desc(mean_catch))
head(mean_region)
```

## Combining datasets together with join()
```{r}
region_defs <- read_csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.303.1") %>% 
  select(code, mgmtArea)

head(catch_long)
head(region_defs)
catch_joined <- left_join(catch_long, region_defs, by = c("Region" = "code"))
#join by columns with different names
```


```{r}
region_defs <- read_csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.303.1") %>% 
  select(code, mgmtArea) %>% 
  rename(Region = code, Region_name=mgmtArea)

catch_joined <- left_join(catch_long, region_defs, by = c("Region"))
```


## split and combine values in columns
```{r}
sites_df <- data.frame(site = c("SJK-323",
                                "JFK-420",
                                "GAF-696"))
head(sites_df)

split_df <- sites_df %>% 
  separate(site, c("island", "site_number"), sep = "-")
head(split_df)

dates_df <- data.frame(
  year = c("1930", "1930", "1932"),
  month = c("12", "12", "12"),
  day = c("14", "15", "16")
)
head(dates_df)

dates_unite <- dates_df %>%
  unite(col = date, 
        year, month, day, 
        sep = "-")
head(dates_unite)
```


```{r}
mean_region <- catch_og %>% 
  select(-All, -notesRegCode) %>% 
  mutate(Chinook = if_else(Chinook == "I",
                           "1",
                           Chinook)) %>% 
  mutate(Chinook = as.numeric(Chinook)) %>% 
  pivot_longer(-c(Region, Year),
               names_to = "species",
               values_to = "catch") %>% 
  mutate(catch = catch *1000) %>% 
  group_by(Region) %>% 
  summarize(mean_catch = mean(catch)) %>% 
  arrange(desc(mean_catch))

head(mean_region)
#18 of 2
```


# save our results with write_csv()
```{r}
library(here)
write_csv(mean_region, here("data/mean_catch_by_region.csv"))
```

